% ── Appendix F: Reproducibility Instructions ──
\section{Reproducibility Instructions}
\label{app:reproducibility}

This appendix provides step-by-step instructions to reproduce all results in this paper. The complete pipeline is designed to run at zero cost using publicly available resources.

\subsection{Prerequisites}

\begin{itemize}
    \item Python 3.11 or later
    \item PyTorch 2.0 or later
    \item Kaggle account (for T4 GPU inference)
    \item OpenRouter API key (for API-based models, total cost $<$ \$3)
    \item Git
\end{itemize}

\subsection{Step-by-Step Reproduction}

\paragraph{Step 1: Clone repository.}
\begin{verbatim}
git clone https://github.com/ssamba1/Scoring-Bias-in-LLM-as-a-Judge
cd Scoring-Bias-in-LLM-as-a-Judge
\end{verbatim}

\paragraph{Step 2: Install dependencies.}
\begin{verbatim}
pip install -r requirements.txt
\end{verbatim}
This installs PyTorch, Transformers, OpenAI Python SDK, NumPy, SciPy, Matplotlib, and other dependencies.

\paragraph{Step 3: Run tests.}
\begin{verbatim}
python tests/test_all.py
\end{verbatim}
This verifies that all dependencies and data files are correctly installed. Expected output: ``All tests passed!''

\paragraph{Step 4: Reproduce T4 family analysis (local inference).}
\begin{verbatim}
# Run inference for 7 base-instruct families on evaluation items
python results_rootcause/run_all_analyses.py
\end{verbatim}
This script:
\begin{enumerate}
    \item Loads each model from Hugging Face (7 base + 7 instruct = 14 models)
    \item Runs inference on the 50 evaluation items with 3 probes $\times$ 3 variants $\times$ 3 repeats
    \item Computes $\Delta$ values, flip rates, and effect sizes
    \item Saves results to \texttt{results\_rootcause/}
\end{enumerate}

\paragraph{Step 5: Reproduce OpenRouter analysis (API inference).}
\begin{verbatim}
# Requires OPENROUTER_API_KEY environment variable
export OPENROUTER_API_KEY="your-key-here"
python results_rootcause/study1_results.json
\end{verbatim}
This runs inference for 22 instruct-tuned models via the OpenRouter API. Expected cost: under \$3 for all models.

\paragraph{Step 6: Generate figures.}
\begin{verbatim}
python paper/generate_figures.py
\end{verbatim}
This generates all figures in \texttt{paper/figures/}.

\paragraph{Step 7: Compile paper.}
\begin{verbatim}
cd paper
latexmk -pdf camera_ready_full.tex
\end{verbatim}
Requires a LaTeX distribution (TeX Live 2023+ recommended). The paper compiles with a single run of \texttt{latexmk}.

\subsection{Docker Reproduction}

For fully containerized reproduction:
\begin{verbatim}
docker build -t scoring-bias .
docker run -e OPENROUTER_API_KEY="your-key" scoring-bias
\end{verbatim}

\subsection{Reproducibility Guarantees}

\begin{itemize}
    \item All random operations use seed 42.
    \item Greedy decoding (temperature 0) ensures deterministic output for local models.
    \item The complete dataset (80 items, 40,500+ judgments) is included in the repository.
    \item A frozen snapshot is archived at Zenodo: \url{https://doi.org/10.5281/zenodo.21361920}.
\end{itemize}
