% Macros filled from the real 13-family (0.1-8B) GPU run (results_scaled.json).
\newcommand{\NFAM}{13}
\newcommand{\MAXB}{8}
% ---- abstract sentence fragments ----
\newcommand{\DECISIVESENTENCE}{instruction tuning sharpens the score distribution (entropy $2.04\!\to\!1.45$ bits, 11/13 families);}
\newcommand{\LINKSENTENCE}{yet more decisive judges are, against intuition, \emph{more} biased---entropy correlates \emph{negatively} with bias ($\rho=-0.41$, $p<10^{-3}$), as does the exact variance term ($\rho=-0.25$);}
\newcommand{\RESULTSENTENCE}{instruction tuning increases scoring bias across all five bias types, not reduces it}
\newcommand{\PREDSENTENCE}{decisiveness even rank-predicts bias out-of-sample ($\rho=0.58$)---with the sign reversed, so the most confident judges are the most biased---while within a single judge it is responsiveness, not confidence, that ranks which nuisances hurt (mean within-judge $\rho=+0.64$).}
\newcommand{\GOLDSENTENCE}{nuisance perturbations collapse a judge's good-vs-bad discrimination (rubric reversal drops accuracy $0.98\!\to\!0$), and instruction tuning does not protect it.}
\newcommand{\PATCHSENTENCE}{patching the instruct representation into the base model transfers the shift in a sharply localized mid-network layer band.}
% ---- results prose ----
\newcommand{\PATCHPROSE}{We test causality directly on Qwen2.5-1.5B (28 layers; rubric-order bias; $n=35$ items with a non-trivial base--instruct score gap). Patching the instruct model's score-position residual into the base model is \emph{inert} at early layers (fraction of items moved toward the instruct score $\approx 0$ for layers 0--9), then jumps in a single layer---from $0.06$ at layer~9 to $0.89$ at layer~10---and reaches $100\%$ from layer~14 onward (Fig.~\ref{fig:patch}). The shift is thus written into a sharp, phase-transition-like band around layer~10: before it the base representation carries no change, and after it the transplanted instruct representation \emph{fully} determines the score. This is an intervention, not a correlation (P3), and localizes \emph{where} instruction tuning installs the effect. The same experiment on Qwen2.5-0.5B ($n=9$ items) replicates the key claim---patching reaches full transfer by the mid-network layers---though with a less sharp onset, so P3 holds across models while the sharpness varies with scale.}
\newcommand{\RESULTSPROSE}{Across the \NFAM{} families, the point estimate of the bias change under instruction tuning is \emph{positive for all five bias types} (Table~\ref{tab:v2main}): score ID ($+0.38$), authority ($+0.23$), and verbosity ($+0.24$) with bootstrap 95\% CIs excluding zero, and rubric order ($+0.23$) and reference answer ($+0.21$) with CIs that include zero individually. A per-item linear mixed-effects model pooling all probes (random intercepts for family and item; $n=13{,}000$) confirms the overall increase is significant (instruct coefficient $+0.16$, $p<10^{-3}$), so the aggregate claim---tuning increases bias---does not rest on any single probe. This is the opposite of what one would hope from a ``better'' judge, and it holds for the two \emph{new} bias types (authority, verbosity) we introduce, so it is not an artifact of the original scoring probes.}
\newcommand{\MECHPROSE}{Why does a more capable judge become more biased? Instruction tuning does make the judge more \emph{decisive}: the entropy of its answer distribution falls from $2.04$ to $1.45$ bits (11/13 families; Fig.~\ref{fig:mech}a). By the naive first-order reading of Proposition~\ref{prop:main}---bias grows with $\sqrt{\mathrm{Var}_\sigma(v)}$---sharper should mean less biased. The data say the reverse: across all \NFAM{} families, five bias types, and both base and instruct checkpoints ($n=130$), entropy correlates \emph{negatively} with bias ($\rho=-0.41$, $p<10^{-3}$; Fig.~\ref{fig:mech}b), and so does the exact variance term $\sqrt{\mathrm{Var}_\sigma(v)}$ ($\rho=-0.25$, $p=0.004$). This is not merely the base/instruct split: the relation holds \emph{within} base-only models ($\rho=-0.25$, $p=0.04$) and \emph{within} instruct-only models ($\rho=-0.33$, $p=0.008$) separately, so decisiveness tracks bias continuously across judges rather than as a two-group difference. Nor is it a capability confound: partialling out model size (rank partial correlation given $\log_{10}$ parameters) leaves the relation essentially unchanged ($\rho=-0.38$, $p<10^{-4}$), and size itself correlates \emph{positively} with bias ($\rho=+0.18$, $p=0.04$)---the direction opposite to what a ``bigger models are sharper and fairer'' confound would require. The relation is strong below 3B ($\rho=-0.51$ and $-0.42$ in the $<$1B and 1--3B bands) and flat in the small $>$3B subsample ($n=30$, $\rho=-0.02$), so the largest judges are where the relation is least resolved. Decisiveness does not confer robustness; it accompanies its loss. Corollary~\ref{cor:diff} explains it: bias factorizes into a decisiveness term and a \emph{responsiveness} term $\|\delta_\pi\|$. We measure the second term directly, as the mean total-variation shift a nuisance induces in the answer distribution. It rises sharply with instruction tuning ($0.15\!\to\!0.26$; $d_z=1.44$, 12/13 families, $p<10^{-3}$; Fig.~\ref{fig:mech}c) and correlates strongly and \emph{positively} with bias ($\rho=+0.82$, $n=130$; Fig.~\ref{fig:mech}d)---far more tightly than decisiveness's $\rho=-0.41$, and equally within base-only ($\rho=0.80$) and instruct-only ($\rho=0.81$) models, so it is a continuous property of judges. The decomposition is thus confirmed by direct measurement of both terms: tuning trims decisiveness a little but inflates responsiveness a lot, and responsiveness is what bias tracks. A confident judge is one that \emph{engages} with the prompt, including its nuisance features. Two cluster-respecting checks sharpen where each term lives. A mixed-effects regression with a family random intercept confirms the between-judge entropy--bias relation (coefficient $-0.46$, $p<10^{-5}$, $n=130$). But \emph{within} a single checkpoint, entropy does not rank which probes bias that judge most (mean within-checkpoint $\rho=-0.05$, n.s.\ across 26 checkpoints)---whereas responsiveness does, and strongly (mean within-checkpoint $\rho=+0.64$, positive in 24/26 checkpoints, Wilcoxon $p<10^{-4}$). The two terms thus operate at exactly the levels Corollary~\ref{cor:diff} assigns them: decisiveness is a judge-level trait that tracks a judge's overall bias level, and responsiveness is the judge$\times$perturbation term that determines \emph{which} nuisance hurts a given judge.}
\newcommand{\CROSSPROSE}{Corollary~\ref{cor:diff} makes a first-order prediction at the level of a single (family, probe) cell: under tuning, $\Delta\log(\text{bias}) \approx \Delta\log\sqrt{\mathrm{Var}_\sigma(v)} + \Delta\log\|\delta_\pi\|$, with both right-hand terms measured independently of the bias itself. Across all 65 cells, the \emph{sign} of the predicted change matches the sign of the actual change in $75\%$ of cells (binomial $p<10^{-4}$), and the predicted magnitude rank-correlates with the actual change ($\rho=0.64$, $p<10^{-6}$). The prediction also captures a majority (12/20) of the cells where bias \emph{decreases}---the cells where the decisiveness drop outweighs the responsiveness rise. The decomposition is thus predictive cell-by-cell, not merely as a pooled correlation: given only how tuning changed a judge's score spread and its distributional response to a nuisance, one can predict in which direction---and roughly how much---its measured bias moved.}
\newcommand{\GENPROSE}{The relation holds within the format group (rubric, score ID; $\rho=-0.55$, $p<10^{-3}$) and, with the same sign and significance, within the content group (reference, authority, verbosity; $\rho=-0.29$, $p=0.009$). The two additional bias types---authority and verbosity, drawn from outside the scoring-bias literature---obey the same negative relation, indicating a general property of these judges rather than a quirk of one probe. It is a statement about how these judges score, not about the specific perturbation. A variant-level decomposition adds one asymmetry worth naming: authority bias is driven twice as strongly by \emph{devaluing} the disclosed novice ($67\%$ of the effect) as by rewarding the disclosed expert ($33\%$)---judges punish framed weakness more than they reward framed authority---whereas the format probes split symmetrically ($48$--$52\%$).}
\newcommand{\PREDPROSE}{With the full \NFAM-family panel (26 checkpoints), control-condition entropy predicts a held-out model's bias out-of-sample: leave-one-family-out cross-validation gives $R^2=0.27$ with rank correlation $\rho=0.58$ between predicted and actual bias ($p=0.004$; Fig.~\ref{fig:predictor}). We report the uncertainty honestly: a family-clustered bootstrap places a wide 95\% CI on the $R^2$ ($[-0.62,\,0.57]$), so with 13 families the \emph{quantitative} predictor is suggestive rather than established; what is stable is the rank signal and its direction. And the direction runs \emph{against} intuition---because the entropy--bias relation is negative, the fit says the \emph{more} confident a judge, the \emph{more} biased. Decisiveness is thus a one-call, label-free screen for bias risk, but an inverted one: confident judges should be trusted less, not more. (On the smaller 9-family subset this relationship is not yet resolved, $R^2<0$; it emerges with the larger panel.)}
\newcommand{\LMMPROSE}{The mixed-effects model (dependent variable: per-item absolute deviation from the control score; random intercepts for family and item) estimates a significant positive instruct effect ($+0.16$, $p<10^{-3}$, $n=13{,}000$), corroborating the family-level tests while accounting for item and family heterogeneity.}
\newcommand{\MITIGPROSE}{The decomposition names the right mitigation. Because bias is driven by responsiveness to the nuisance dimension, the fix is to remove that dimension, not to make the judge more confident. Marginalizing the score over the three score-ID formats cuts the measured bias from $\Delta=1.09$ (single-format expected value) to $0.45$---a 59\% reduction---whereas taking the discrete argmax, a \emph{more} decisive readout, \emph{increases} it to $1.88$. A second nuisance-removal intervention works the same way: ensembling the score over three prompt templates cuts single-template bias by 22\% ($0.67\!\to\!0.53$ across 30 model$\times$probe cells, Wilcoxon $p<10^{-5}$), without needing to know which nuisance is present. Confidence-increasing interventions make bias worse; averaging over nuisances makes it better.}
\newcommand{\GOLDPROSE}{With no perturbation, both base and instruct judges discriminate almost perfectly (accuracy $0.98$; Table~\ref{tab:gold}), and instruct judges are the more confident, with a larger good$-$bad margin ($1.26$ vs $0.74$)---consistent with their sharper distributions. Nuisances then do real damage: rubric reversal collapses accuracy to chance for both ($0.02$ base, $0.00$ instruct), and verbosity padding costs about 30 points ($0.70$/$0.73$); authority framing is nearly harmless ($0.97$/$0.97$). Instruction tuning is \emph{not} protective---its margin loss equals or exceeds the base model's (reversed $2.43$ vs $1.33$; verbose $0.97$ vs $0.71$)---again the pattern that a more decisive judge, having committed harder, has more to lose. Bias is not cosmetic: it corrupts real quality judgments (P6).}
\newcommand{\SECONDARYPROSE}{Flip rates (the fraction of items whose discrete score changes under the nuisance, comparable to \citet{li2025scoring}) are substantial for both checkpoints---e.g.\ $0.75$/$0.71$ (base/instruct) for score ID---and rise for the content probes after tuning (reference answer $0.22\!\to\!0.38$; authority $0.24\!\to\!0.41$), mirroring the $\Delta$ increases. The effect is not domain-specific: instruct bias exceeds base bias in every one of the five item domains. Across the $0.1$--$\MAXB$B range we tested the increase persists at the largest (7--8B) models, so it is not confined to the smallest judges; a preregistered 4-bit Qwen2.5-14B extension (P12) is nominally positive but strongly attenuated ($+0.06$ vs the panel's $+0.26$; positive for 3/5 probes), consistent with the $>$3B attenuation of the entropy--bias relation and subject to a quantization caveat. Nor is it driven by any single family or vendor: leaving out each family in turn keeps the mean instruct$-$base change within $[+0.23, +0.29]$; excluding all four Qwen families gives $+0.26$ (positive in 8/9 remaining families); and restricting to $\geq$1B models gives $+0.29$ (9/10). Finally, the expected-value and discrete readouts agree (Spearman $\rho=0.56$ between EV bias and flip rate across the 130 model$\times$probe cells), so the finding is not an artifact of the readout.}
\newcommand{\ATTNPROSE}{The natural hypothesis---that tuning raises responsiveness by making the judge attend \emph{more} to the injected nuisance---is \emph{refuted}. Across three Qwen families, instruct judges attend \emph{equally or slightly less} to the nuisance span than base judges (instruct/base attention ratio $0.95$--$1.00$ in all six model$\times$perturbation cases). Responsiveness therefore rises \emph{despite} flat input attention: the same attended content becomes more score-consequential \emph{downstream}, consistent with the patching result that the change lives in a mid-network computation (\S\ref{sec:patch}), not in attention allocation. We report this null deliberately: a prior, fabricated version of this project invented a positive attention ``mechanism'' at exactly this point; the real measurement does not support the simplest such account, and identifying the precise circuit is left to future work.}
\newcommand{\MTPROSE}{The negative decisiveness--bias relation is not an artifact of one prompt format: it holds under all three templates (Qwen2.5-0.5B/1.5B/3B, $n=30$ points each), with $\rho=-0.46$ ($p=0.011$), $-0.43$ ($p=0.016$), and $-0.63$ ($p<10^{-3}$). The effect is a property of these judges, not of a particular way of asking.}
\newcommand{\SPECPROSE}{We stress-test the headline across twelve analysis specifications: $\{$expected-value, argmax$\}$ readout $\times$ $\{$max--min spread, mean $|$deviation$|\}$ metric $\times$ $\{$all, format, content$\}$ probe sets. All six expected-value specifications give a positive mean effect ($9$--$11$/13 families positive), and the effect passes an exact sign-flip permutation test enumerating all $2^{13}$ family sign patterns ($p=0.00098$; per-family forest with probe-resampled CIs in Fig.~\ref{fig:forest}). The discrete-argmax specifications are the honest outliers: quantizing each score to its modal value washes the increase out for the format probes ($-0.15$ to $-0.07$) and preserves it only for the content probes ($+0.03$ to $+0.11$)---consistent with the flip-rate pattern (\S\ref{sec:secondary}), where content flips rise after tuning but format flips do not. The increase is thus a property of the judge's \emph{continuous} scoring distribution, which the expected-value readout exposes; the argmax readout simultaneously \emph{inflates} the level of measured bias (1.88 vs.\ 1.09 in \S\ref{sec:lmm}) and \emph{compresses} its change under tuning---quantization exaggerates and hides at once. Practitioners reading discrete scores will underestimate how much tuning has raised their judge's bias. Measurement noise is not at issue: the $\Delta$ estimator's split-half reliability is $0.99$ (Spearman--Brown, 130 cells), so all of these are stable properties of the judges.}
\newcommand{\DOSEPROSE}{Does bias scale with the \emph{magnitude} of the nuisance? We preregistered (P14) that it would---graded verbosity padding (0/1/2/4/8 filler units) and five graded authority framings, four families, base and instruct---and that instruct dose--response slopes would be steeper. \emph{Both clauses failed}: the mean dose--rank correlation of the shift is $0.06$ (8/16 cells positive), and instruct slopes are not steeper (3/8 pairs, $p=0.74$). The observed shape is a step function: most of the shift appears at the first dose unit and then plateaus (e.g., Qwen2.5-1.5B base authority: $0\!\to\!0.61$ at one unit, then \emph{declining} with stronger framing). Responsiveness is triggered by the nuisance's \emph{presence}, not scaled by its magnitude---consistent with the span-patching result that the authority channel is a (categorical) property of how the framing sentence is encoded (\S\ref{sec:spanpatch}). The decomposition itself is unaffected ($\|\delta_\pi\|$ was never specified as dose-proportional), but our auxiliary intuition that it would be is refuted, and we report it as such.}
\newcommand{\SPANPROSE}{A second, sharper causal test asks \emph{where the nuisance itself is encoded} (preregistered P13). We run the instruct model on the perturbed prompt but overwrite its residual stream \emph{at the nuisance-span token positions} with the base model's activations for the same prompt, layer by layer (Qwen2.5-1.5B, $n=20$). For the \textbf{authority} framing the result is decisive: the patch erases $\geq$50\% of the instruct-vs-base response gap across layers 3--14, peaking at $\approx$100\% at layers 6--11---given the base model's encoding of the framing sentence, the instruct model responds like the base model. Combined with the flat attention result (\S\ref{sec:attn}), this localizes the authority channel precisely: tuning changes \emph{what is written into} the nuisance span's representation in the early-to-mid layers, not how much the score position attends to it. For the \textbf{good-exemplar} framing the same patch does nothing at any layer (max reduction $7\%$): exemplar responsiveness is not span-local, and its mechanism is distributed---a preregistered clause that failed, reported as such. Content nuisances are thus not mechanistically uniform, which any single-mechanism account of judge bias would miss.}
\newcommand{\NEWPROBEPROSE}{Two further content perturbations, preregistered as P10 with a committed analyzer before the run finished, probe how far the pattern extends. \textbf{Sycophancy} (the user states an opinion---``I think this response is excellent/quite poor''---before the scoring request) shows the \emph{largest tuning effect in this entire study}: instruct $\Delta$ exceeds base $\Delta$ in 11/13 families, mean change $+0.46$ (Wilcoxon $p=0.027$), with an instruct-side bias of $1.37$---larger than any of the original five probes. This is the judge-side face of the sycophancy documented in assistants, and it is \emph{installed} by the same tuning. \textbf{Anchoring} (a numeric prior: ``most responses of this kind receive a score of 5/1'') is the preregistered prediction that \emph{failed}: the mean change is $+0.015$ ($p=0.54$), and a formal equivalence test bounds it---the family-bootstrap 90\% CI $[-0.13, 0.14]$ lies within a $\pm 0.15$ margin, so the null is affirmed, not merely unrejected. The contrast is informative: tuning inflates responsiveness to \emph{social and content framing}, not to every injected signal---a numeric anchor, which a socially-tuned judge has no learned reason to defer to, leaves bias unchanged. The decomposition accommodates both outcomes ($\|\delta_\pi\|$ is perturbation-specific), but did not predict which perturbations would move; that remains an empirical fact about what instruction tuning teaches.}
\newcommand{\ANATPROSE}{Where does the variance in bias live? An ANOVA decomposition of the 130 (family, checkpoint, probe) bias cells attributes $8\%$ of the variance to the judge family, $24\%$ to the probe, and $6\%$ to the base-vs-instruct distinction---but $37\%$, the single largest share, to the family$\times$probe \emph{interaction}. Bias is chiefly a property of the judge--perturbation \emph{pair}, exactly the signature Corollary~\ref{cor:diff} predicts for a responsiveness-dominated regime. Consistent with this, items carry essentially no stable bias signal: item-level bias profiles correlate near zero across judges (mean $\rho=0.01$ over $1{,}625$ same-probe pairs)---no item is ``hard'' in itself. Finally, the sharpening that tuning produces is genuine concentration, not leniency: the maximum answer-token probability rises $0.40\!\to\!0.53$ while the mass on the top scores (4--5) is flat ($0.54\!\to\!0.51$), ruling out the deflationary reading that tuned judges simply score higher. Nor is bias a scalar judge trait: a judge's bias on one probe barely predicts its bias on another (mean cross-probe Spearman $0.16$ across the ten probe pairs), with only the content pair authority--verbosity appreciably correlated ($0.60$)---again the judge$\times$perturbation structure, now visible in the trait geometry itself.}
\newcommand{\STAGEPROSE}{``Instruction tuning'' bundles several stages; which one installs the effect? We run the full alignment ladder---base $\to$ SFT $\to$ DPO $\to$ RLVR---for OLMo-2 (1B, 7B) and Tülu-3-8B, whose intermediate checkpoints are public, with predictions preregistered before the run finished (\texttt{PREREGISTRATION.md}, P7--P9). \textbf{SFT installs the responsiveness} (P7, confirmed): responsiveness rises from base to SFT in 10/10 family$\times$probe cells, and the SFT step alone accounts for $84\%$ (OLMo-1B) and $99\%$ (OLMo-7B) of the total base$\to$RLVR responsiveness rise. Bias jumps with it (OLMo-1B $\Delta$: $0.24\!\to\!0.80$ at SFT). \textbf{Preference tuning installs the confidence} (P8, confirmed): DPO and RLVR keep sharpening the distribution (OLMo-1B entropy $2.21\!\to\!1.80\!\to\!1.04\!\to\!0.99$ bits) while adding little or no further responsiveness (OLMo-7B: DPO $-0.01$, RLVR $+0.01$, vs.\ SFT's $+0.06$). Across the eight stage transitions, bias moves with responsiveness in 7/8 (P9; binomial $p=0.07$, marginal at this $n$---reported as preregistered). The paper's title phenomenon thus decomposes cleanly across the pipeline: supervised fine-tuning teaches the judge to engage with the prompt---including its nuisances---and preference optimization then teaches it to be sure. One boundary is worth stating: \emph{across} the 11 ladder checkpoints the entropy--bias correlation is absent ($\rho=0.01$, $n=55$)---like the within-checkpoint and $>$3B analyses, this scopes the negative decisiveness--bias relation as a \emph{between-family} regularity of the full panel, not a law visible inside every subgroup.}
\newcommand{\REPLPROSE}{Our 50 evaluation items are author-written; to remove item authorship as a variable we replicate on 50 items drawn deterministically (seed 42) from the public \emph{databricks-dolly-15k} dataset (open\_qa/general\_qa, responses truncated to one sentence to hold quality mid-scale), across the eight $\leq$3B families. Both headline findings replicate: instruction tuning increases mean bias in 7/8 families (mean $+0.13$; positive for four of five bias types, with rubric order flat at $-0.04$), and the entropy--bias relation reproduces almost exactly ($\rho=-0.44$, $p<10^{-4}$, $n=80$, vs.\ $-0.41$ on the original items). The effect is smaller on these shorter, factoid-style items, consistent with content nuisances having less surface to act on, but its direction and the core negative relation are item-set independent. Synthesizing across the three datasets now available---original items, public items, and the three-template rerun---the tuning effect on bias is positive in 26 of 30 family-level measurements, with an $n$-weighted mean of $+0.23$ (a descriptive synthesis; the datasets share model families).}
\newcommand{\ZHPROSE}{The findings are not English-specific. On a fully Chinese version of the suite---all 50 items, rubrics, labels (native equivalents of Poor--Excellent), and framings translated; Qwen2.5 0.5--7B, natively strong in Chinese; preregistered as P11 with the analyzer committed first---both headline findings replicate: instruction tuning increases mean bias in 4/4 families (mean $+0.43$; positive point estimate for every probe), and the entropy--bias correlation is again negative ($\rho=-0.36$, $p=0.024$, $n=40$).}
\newcommand{\DISCUSSION}{Usefully, the audit is cheap: one unperturbed scoring call, whose entropy is an inverted proxy for bias risk---trust the hedging judge over the confident one.}
\newcommand{\CONCLUSION}{Across \NFAM{} open model families spanning $0.1$--$\MAXB$B and five bias types, instruction tuning makes LLM judges \emph{sharper yet more biased}: it lowers the entropy of the score distribution while raising bias across all five bias types. This refutes the intuitive expectation that a more decisive judge is a more robust one---empirically, decisiveness correlates negatively with bias, and even predicts it out-of-sample with the sign reversed. Our theory reconciles the two: bias factorizes into a decisiveness term and a responsiveness term (Proposition~\ref{prop:main}, Corollary~\ref{cor:diff}), and instruction tuning inflates responsiveness faster than it trims decisiveness. Causal patching localizes where this is written into the network; a ground-truth test shows the bias corrupts real good-vs-bad judgments; and a simple marginalization mitigates it where confidence-raising cannot. Confidence is not robustness. We release a single-script reproduction and a full integrity audit of a prior, fabricated version of this project.}
