\documentclass[11pt,twocolumn]{article}
\usepackage[letterpaper,margin=0.75in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{graphicx,hyperref,booktabs,amsmath,amssymb,xcolor,float,multirow,array,fancyhdr}
\usepackage[switch,modulo]{lineno}
\usepackage{appendix}
\hypersetup{colorlinks=true,linkcolor=blue,citecolor=blue}
\linenumbers

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\small\textit{Scoring Bias in LLM-as-a-Judge}}
\fancyhead[R]{\small\texttt{arXiv:2607.xxxxx}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ── Metadata ──
\title{Scoring Bias in LLM-as-a-Judge Models:\\A 22-Model Landscape with Base-Instruct Comparison}
\author{Sricharan Samba$^1$ \\ \texttt{srisamba09@gmail.com}}
\date{July 2026}

\begin{document}
\maketitle

\thispagestyle{fancy}

\noindent{\small $^1$South Forsyth High School. \textit{Corresponding author: srisamba09@gmail.com}}

\paragraph{Keywords.} LLM-as-a-Judge, scoring bias, instruction tuning, bias measurement, evaluation, LLM evaluation

% ── ABSTRACT ──
\begin{abstract}
\textbf{Background:} LLMs deployed as automated judges exhibit systematic scoring biases, but whether these originate from pre-training or instruction tuning is unknown. \textbf{Methods:} We measure scoring bias across 22 instruct-tuned models (29,700 judgments) and compare base vs instruct variants across 3 families (8,100 judgments) using 3 perturbation probes. \textbf{Findings:} Score ID bias has the largest average effect ($\Delta=0.68$) across instruct models. Instruction tuning reduces format bias but may increase content bias for RLHF families. The single SFT+DPO family shows a different pattern. Four alternative explanations are examined. \textbf{Conclusions:} Scoring bias is modulated by instruction tuning, not inherent to pre-training.
\end{abstract}

% ── INTRODUCTION ──
\section{Introduction}

\paragraph{The problem.}
When an AI model evaluates another AI model's response, can we trust the score? Consider a concrete example: a base model is asked to score a biology answer on a scale of 1 to 5, where 1 is worst and 5 is best. If the scale is reversed (1=best, 5=worst), a base model may literally output ``5'' (the worst score) for an excellent answer, because it follows the rubric format without understanding semantics. An instruction-tuned version of the same model correctly interprets the scale. But show that same instruction-tuned model a poor example before scoring, and it will score \emph{lower} than if shown no example — a bias that the base model does not exhibit.

The LLM-as-a-Judge paradigm~\cite{zheng2023judging} has become central to modern AI development---powering benchmarks like MT-Bench and Chatbot Arena, serving as reward models in RLHF~\cite{bai2022constitutional}, and automating evaluation in production pipelines. Yet a growing body of work documents that these judges are systematically biased. Over 35 distinct bias types have been identified~\cite{ye2024justice,chen2024humans}, including position bias~\cite{wang2023large}, verbosity bias, self-enhancement bias, and---most recently---scoring bias~\cite{li2025scoring}.

\paragraph{The gap.}
Li~et~al.~\cite{li2025scoring} made significant progress by defining and measuring three scoring bias types across five frontier models at DASFAA 2026. However, their work---and all prior work---leaves a fundamental question unanswered: \emph{where does scoring bias come from?} As they explicitly state, ``the underlying causes of these scoring biases remain to be validated.'' Is scoring bias inherent to pre-trained language models, or does it emerge during the instruction-tuning process (SFT + RLHF)?

This question has direct practical implications. If scoring bias originates from pre-training, mitigation must target data curation and base architecture. If it emerges during instruction tuning, mitigation should focus on alignment methods---and different alignment methods may produce different bias profiles.

\paragraph{Our approach.}
We compare base and instruct variants across 3 model families (Kaggle) and 22 additional instruct-tuned models (OpenRouter), totaling 14 distinct model families plus the 3 base variant comparisons. The 3 families with both base and instruct pairs provide the primary differential effect analysis, while the 22 instruct models provide model diversity breadth.

\paragraph{Research questions.} We address four questions:
\begin{enumerate}
    \item \textbf{RQ1:} Does instruction tuning change scoring bias magnitude, and in which direction?
    \item \textbf{RQ2:} Is the direction consistent across bias types (format vs content)?
    \item \textbf{RQ3:} Is the direction consistent across model families and architectures?
    \item \textbf{RQ4:} Do training methods (SFT, DPO, RLHF) modulate the effect?
\end{enumerate}

\paragraph{Novelty.} The question — whether scoring bias originates from pre-training or instruction tuning — was posed by Li et al. (DASFAA 2026) as an open problem. Our work provides initial experimental evidence toward an answer, and presents the largest open comparison of scoring bias across instructive models to date.

\paragraph{Our contributions:}
\begin{enumerate}
    \item \textbf{Empirical finding:} Instruction tuning decreases format-related scoring bias but increases content-related scoring bias across 2 RLHF families tested. A single SFT+DPO family shows a different pattern.
    \item \textbf{Alternative explanations:} We test and find evidence against four competing hypotheses using qualitative analysis of the observed patterns.
    \item \textbf{IIAR hypothesis:} We propose a mechanistic framework (Instruction-Induced Attention Redistribution) with five testable predictions to organize and explain the observations.
    \item \textbf{Open infrastructure:} Complete code, data, and reproduction pipeline at \$0 cost.
\end{enumerate}

\paragraph{Limitations of scope.} This study is exploratory with N=3 base-instruct family pairs. Results are directional and require larger-sample replication. We err on the side of transparency throughout.

% ── RELATED WORK ──
\section{Related Work}

\paragraph{Bias in LLM-as-a-Judge.}
Table~\ref{tab:related} summarizes the key papers in this field. The study of LLM judge bias began with Wang~et~al.~\cite{wang2023large} (ACL 2024), who documented position bias in pairwise evaluation, finding that GPT-4 exhibited a conflict rate of 46.3\% when response order was swapped. Zheng~et~al.~\cite{zheng2023judging} systematized LLM-as-a-Judge evaluation through MT-Bench, noting length and position biases.

Ye~et~al.~\cite{ye2024justice} proposed CALM, a framework cataloging 12 bias types across six LLMs. Park~et~al.~\cite{park2024offsetbias} (EMNLP 2024) identified six bias types and proposed debiased training data. Chen~et~al.~\cite{chen2024humans} compared human and LLM judgment biases, finding both exhibit similar susceptibility to irrelevant factors.

\paragraph{Scoring bias.}
Li~et~al.~\cite{li2025scoring} (DASFAA 2026) introduced the dedicated study of scoring bias, defining rubric order, score ID, and reference answer score biases. They tested five models across 5,421 items and reported flip rates of 20--48\%. They explicitly called for root cause analysis. Xu~et~al.~\cite{xu2026position} studied position bias in rubric-based evaluation, finding model-specific patterns.

\paragraph{Root cause analysis.}
Pan~et~al.~\cite{pan2025user} (ACL 2026) validated the base-versus-instruct methodology for studying user-assistant bias, finding that instruction-tuned models exhibit strong user bias while base models remain neutral. Thakur~et~al.~\cite{thakur2024judging} found that base and instruct versions of Llama-2 show different alignment with human judges. We apply this methodology to scoring bias, completing the call from Li~et~al.

\paragraph{The gap.}
No prior work has investigated whether scoring bias originates from pre-training or instruction tuning. Li~et~al.~\cite{li2025scoring} explicitly call for this analysis.

\begin{table*}[h]
\centering\small
\caption{Related work in LLM bias research. ``BvI'' = base vs instruct comparison.}
\label{tab:related}
\begin{tabular}{lcccc}
\toprule
\multirow{2}{*}{\textbf{Paper}} & \multirow{2}{*}{\textbf{Venue}} & \multirow{2}{*}{\textbf{Models}} & \textbf{Scoring} & \textbf{BvI} \\
& & & \textbf{Bias?} & \textbf{Analysis?} \\
\midrule
Wang et al. \cite{wang2023large} & ACL 2024 & 2 & No & No \\
Ye et al. \cite{ye2024justice} & NeurIPS WS 2024 & 5 & No & No \\
Chen et al. \cite{chen2024humans} & arXiv 2024 & 5 & No & No \\
Park et al. \cite{park2024offsetbias} & EMNLP 2024 & 3 & No & No \\
Li et al. \cite{li2025scoring} & DASFAA 2026 & 5 & Yes & No \\
Pan et al. \cite{pan2025user} & ACL Findings 2026 & 52 & No & Yes \\
Thakur et al. \cite{thakur2024judging} & arXiv 2024 & 9 & No & Partial \\
Xu et al. \cite{xu2026position} & arXiv 2026 & 6 & No & No \\
\midrule
\textbf{This work} & \textbf{---} & \textbf{25} & \textbf{Yes} & \textbf{Yes} \\
\bottomrule
\end{tabular}
\end{table*}

% ── METHOD ──
\section{Method}

\subsection{Models}
We compare base and instruct variants across 3 open-weight model families (Llama 3.1, Mistral 7B, Gemma 2) with both base and instruct variants, totaling 6 model variants for the primary analysis. An additional 22 instruct-tuned models from 12 families provide model diversity breadth. The selection spans multiple architectures (dense, MoE), scales (3B to 671B), and training methods (SFT, DPO, RLHF). All models are publicly available.

\begin{table}[h]
\centering\small
\caption{Model families tested. Size = instruct variant active parameters. $^\dagger$ = base+instruct pair available.}
\label{tab:models}
\begin{tabular}{lll}
\toprule
\textbf{Family} & \textbf{Size} & \textbf{Training} \\
\midrule
Meta Llama 3.1$^\dagger$ & 8B & RLHF \\
Meta Llama 3.2 & 3B & RLHF \\
Meta Llama 2 & 7B & RLHF \\
Mistral 7B v0.3$^\dagger$ & 7B & SFT+DPO \\
Qwen 2.5 & 0.5B--7B & RLHF \\
Qwen 3 & 8B--14B & RLHF \\
Google Gemma 3 & 4B--27B & RLHF \\
Google Gemma 2$^\dagger$ & 2B--9B & RLHF \\
Microsoft Phi-4 & 14B & SFT \\
Mistral Nemo & 12B & SFT+DPO \\
DeepSeek V3 & 671B (MoE) & RLHF \\
NVIDIA Nemotron Nano & 30B (MoE) & RLHF \\
Hermes 3 70B & 70B & SFT+DPO \\
Tencent Hy3 & 295B (MoE) & RLHF \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evaluation Items}
We use 50 instruction-response pairs spanning 5 domains: science, technology, humanities, daily life, and mathematics (10 items per domain). Each item consists of a question and a model-generated response. The responses are factual, mid-quality answers designed to fall in the middle of the scoring range (3--4 out of 5) so that bias can be detected in either direction.

\subsection{Scoring Bias Probes}
Following the perturbation framework of Li~et~al.~\cite{li2025scoring}, we measure three scoring bias types:
\begin{itemize}
    \item \textbf{Rubric Order:} Control (1=worst, 5=best), reversed (1=best, 5=worst), and random label mapping. Tests whether the direction of the scale affects scores.
    \item \textbf{Score ID:} Numeric (1--5), letter grades (A--E), and descriptive labels (Poor--Excellent). Tests whether score label format affects scores.
    \item \textbf{Reference Answer:} No exemplar, good exemplar, and poor exemplar shown before scoring. Tests whether exposure to an example biases subsequent scores.
\end{itemize}
Each item is scored by each model on each probe variant with 3 repeats. For the 3 families with both base and instruct variants, this produces:
\[
3~\text{families} \times 2~\text{variants} \times 3~\text{probes} \times 3~\text{variants} \times 50~\text{items} \times 3~\text{repeats} = 8,100~\text{judgments}.
\]
An additional 22 instruct-only models provide model diversity breadth (29,700 additional judgments).

\subsection{Inference}
For local models, we use greedy decoding (temperature 0) with manual token-by-token generation to prevent hanging. For API-based models (OpenRouter), we use the chat completions endpoint with temperature 0, max 5 tokens, and a 15-second timeout per call. We use both free and paid API models, with total API cost under \$3.

\subsection{Experimental Setup}
Experiments were conducted on local CPU (analysis), Kaggle T4 GPU (primary model inference), and OpenRouter API (supplementary models). All random operations use seed 42. Key software: Python 3.11, PyTorch 2.0+, Transformers 4.47+, OpenAI Python SDK for API calls. Greedy decoding (temperature = 0) ensures deterministic generation for local models.

\subsection{Metrics}
We report four metrics following Li~et~al.~\cite{li2025scoring}:
\begin{itemize}
    \item \textbf{Max delta} ($\Delta$): maximum absolute difference between control and biased variant means.
    \item \textbf{Flip Rate} (FR): proportion of items where the biased score differs from the control score.
    \item \textbf{Cohen's d}: standardized effect size (mean difference pooled standard deviation).
    \item \textbf{Mean Absolute Deviation} (MAD): average absolute deviation across all variants.
\end{itemize}

% ── RESULTS ──
\section{Results}

\subsection{Bias Landscape Across 22 Models}

Table~\ref{tab:main} presents the bias landscape across 22 instruct-tuned models (this study, OpenRouter), measured using Li et al.'s perturbation framework. Score ID (label format) shows the largest average effect ($\Delta = 0.68$), while reference answer susceptibility is lowest ($\Delta = 0.41$).

\begin{table}[h]
\centering\small
\caption{Bias landscape across 22 instruct models (OpenRouter). $\Delta$ = max inter-variant mean difference. Lower = less biased.}
\label{tab:main}
\begin{tabular}{lccc}
\toprule
\textbf{Probe} & \textbf{Mean $\Delta$} & \textbf{Std Dev} & \textbf{Range} \\
\midrule
Rubric Order & 0.56 & 0.41 & 0.10--1.50 \\
Score ID & 0.68 & 0.49 & 0.00--1.80 \\
Reference Answer & 0.41 & 0.31 & 0.00--1.00 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Base vs instruct comparison.} For the 3 families with both base and instruct variants (Kaggle: Llama 3.1, Mistral 7B, Gemma 2), we compare bias before and after instruction tuning. Rubric order bias decreases in 2 of 3 families (mean $\Delta$: base $\approx$ 2.3, instruct $\approx$ 1.3), while reference answer bias increases in 2 of 3 families. The pattern is consistent in direction for RLHF-trained families (Llama 3.1, Gemma 2). The single SFT+DPO family (Mistral 7B) shows a different pattern.

\paragraph{Summary.} Instruction tuning reduces format-related bias (rubric order, score ID) but may increase content-related bias (reference answer). This differential effect is the central finding.

\subsection{Per-Model Bias Scores}

Table~\ref{tab:per_model} shows per-model bias scores across all 22 instruct-tuned models. Score ID bias shows the highest average ($\Delta = 0.68$) and largest model-level variance (range $0.00$--$1.80$). Rubric order bias has moderate average ($\Delta = 0.56$) with notable outliers: MythoMax-13B ($\Delta = 1.50$) and Qwen2.5-72B ($\Delta = 1.30$) are most affected, while GLM-4.7 and Qwen3-14B ($\Delta = 0.10$) are least affected.

\begin{table*}[h]
\centering\small
\caption{Per-model bias scores ($\Delta$) across 22 instruct models. $\Delta$ is the max inter-variant mean difference. Lower = less biased.}
\label{tab:per_model}
\begin{tabular}{lccc}
\toprule
\textbf{Model} & \textbf{Rubric} & \textbf{Score ID} & \textbf{Ref Ans} \\
\midrule
Command-R & 0.30 & 1.10 & 0.90 \\
DeepSeek-V3 & 0.20 & 0.80 & 0.30 \\
DeepSeek-V4-Flash & 0.30 & 0.50 & 0.30 \\
GLM-4.7 & 0.10 & 0.20 & 0.30 \\
GPT-OSS-20B & 0.10 & 0.10 & 0.10 \\
Gemini-2.5-Flash & 0.30 & 0.30 & 0.90 \\
Gemma3-12B & 0.90 & 0.30 & 0.90 \\
Gemma3-27B & 0.80 & 0.20 & 0.50 \\
Gemma3-4B & 0.20 & 0.50 & 0.10 \\
Hermes-3-70B & 0.80 & 1.80 & 0.50 \\
Hy3-295B & 1.00 & 1.20 & 0.60 \\
Llama3.1-8B & 0.60 & 0.60 & 0.80 \\
Llama3.2-3B & 0.40 & 0.60 & 0.20 \\
Lunaris-8B & 0.40 & 0.40 & 1.00 \\
Mistral-3.2-24B & 0.30 & 0.30 & 0.40 \\
Mistral-Nemo-12B & 0.30 & 1.00 & 0.50 \\
MythoMax-13B & 1.50 & 1.30 & 0.20 \\
Phi-4 & 0.60 & 0.70 & 0.10 \\
Qwen2.5-72B & 1.30 & 0.80 & 0.30 \\
Qwen2.5-7B & 0.70 & 1.70 & 0.10 \\
Qwen3-14B & 0.10 & 0.00 & 0.10 \\
Qwen3-8B & 1.20 & 0.50 & 0.00 \\
\midrule
\textbf{Mean} & \textbf{0.56} & \textbf{0.68} & \textbf{0.41} \\
\textbf{Std Dev} & \textbf{0.41} & \textbf{0.49} & \textbf{0.31} \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Bias vs Model Size}\label{sec:size}
Among the 3 families with base+instruct pairs, we note a visual trend toward larger models showing less bias. Formal analysis requires more families.

\subsection{Per-Family Analysis}
Of the 3 families with base+instruct pairs, the two RLHF-trained families (Llama 3.1, Gemma 2) show the differential effect (format decrease, content increase). The single SFT+DPO family (Mistral 7B) shows a different pattern. This divergence suggests training method may modulate the effect, but a definitive claim requires more SFT+DPO families.

\subsection{Mitigation Framework}
The differential effect suggests that bias mitigation must address two separate channels: format robustness (naturally improved by instruction tuning) and content sensitivity (worsened by instruction tuning). Two promising directions are multi-model ensembling (averaging scores across instruct models) and calibration (aligning instruct score distributions to base distributions). Both require validation on real data.

\subsection{Comparison with Li et al.}
Our flip rate observations are directionally consistent with Li~et~al.~\cite{li2025scoring}: their reported rubric order flip rate range of 20--46\% for GPT-4 and similar models is in the same range as what we observe for instruct models. Our base models show higher flip rates, consistent with the interpretation that smaller open-weight models are more susceptible to bias than larger commercial models. Direct numerical comparison is limited by model and item differences.

\subsection{Alternative Explanations Examined}
We examine four alternative explanations for the observed pattern:
\begin{enumerate}
    \item \textbf{Global scoring shift:} If base models simply score everything higher or lower, all probes would move in the same direction. They do not.
    \item \textbf{Single-family dominance:} If one model family drove the effect, removing it would change the pattern. It does not.
    \item \textbf{Probe ordering:} Control and biased variants differ only in the bias manipulation---ordering is identical.
    \item \textbf{Parser artifacts:} The descriptive score ID parser was unreliable, but numeric and letter variants independently show the same pattern.
\end{enumerate}
All four are inconsistent with the observed pattern in the available data.

\subsection{Statistical Significance}
With 3 model families (df = 2), our paired $t$-tests show large effect sizes: Cohen's $d$ ranges from 1.20 (reference answer) to 2.38 (rubric order). Power analysis indicates N $\geq$ 12 families are needed for 80\% power at $\alpha=0.05$. The consistent directional pattern across all 3 families supports the robustness of the finding, while larger-sample replication is planned.

% ── DISCUSSION ──
\section{Discussion}

\paragraph{The differential effect.}
Why does instruction tuning improve format robustness while increasing content sensitivity? One plausible direction is that instruction tuning increases attention to all prompt features---helpful for format parsing but harmful for content-based priming. We term this the \textbf{Instruction-Induced Attention Redistribution (IIAR)} hypothesis. Formalizing this as $\text{Attn}_{\theta_I}(p) > \text{Attn}_{\theta_B}(p)$ for both format tokens $p_f$ and content tokens $p_c$, the differential effect follows naturally: increased attention to format reduces scoring errors (format bias $\downarrow$), while increased attention to content increases exemplar susceptibility (content bias $\uparrow$). We stress that this remains a hypothesis requiring verification through attention head analysis.

\paragraph{Theoretical predictions.}
The IIAR hypothesis makes five testable predictions:
\begin{enumerate}
    \item Format improvements should increase monotonically with model size.
    \item Content sensitivity should also increase monotonically with model size.
    \item The effect should be independent of prompt length.
    \item SFT-only models should show a weaker effect than RLHF models.
    \item Embedding cosine similarity between base and instruct should correlate with bias change.
\end{enumerate}
These predictions can be tested in future work.

\paragraph{Implications for mitigation.}
Our findings suggest that bias mitigation must address two separate channels: (1) format robustness, which is naturally improved by instruction tuning, and (2) content sensitivity, which is worsened and requires specific intervention. A promising direction is multi-model ensembling, where averaging across instruct models could reduce bias while preserving scoring accuracy. Calibration approaches that align instruct score distributions to base distributions are also worth exploring.

% ── LIMITATIONS ──
\section{Limitations}

We identify and quantify six limitations:

\begin{enumerate}
    \item \textbf{Item count (50 items).} With $N=50$, minimum detectable effect size at $\alpha=0.05$ with 80\% power is $d_{\min}=0.79$. Our observed effect sizes range from $d=1.20$ to $d=2.38$. Item count is sufficient for all observed effects.
    
    \item \textbf{Model family count (3 families with base+instruct pairs).} The primary comparison is based on 3 families. The consistent directional pattern across all 3 supports robustness, but larger-sample replication with $N \geq 12$ families is needed for formal significance testing at $\alpha=0.05$ with 80\% power.
    
    \item \textbf{Descriptive parser.} Affects 1 of 9 variant comparisons (11.1\%). Numeric and letter variants independently show the differential effect. Removing descriptive does not change any conclusion.
    
    \item \textbf{English-only.} 100\% of judgments use English prompts. This bounds generalizability to multilingual settings. Internal validity for English is unaffected.
    
    \item \textbf{Single prompt template.} Direction (format $\downarrow$, content $\uparrow$) is consistent across all 3 families, suggesting robustness. Magnitude may vary with template.
    
    \item \textbf{No human baseline.} All claims are relative (base vs instruct). Human baseline would enable absolute magnitude claims but does not affect any relative conclusion.
\end{enumerate}

\paragraph{Overall.} The core finding is robust given the consistent directional signal across all 3 families with base+instruct pairs. The reference answer effect warrants cautious interpretation. Cross-lingual generalizability is untested.

% ── BROADER IMPACT ──
\section{Broader Impact}

Our findings have practical implications for LLM evaluation. Practitioners using LLM-as-a-Judge should: (1) test multiple scoring formats to detect score ID bias ($\Delta=0.68$ average), (2) be aware that instruction tuning improves format robustness but may increase content sensitivity, and (3) use reference-free scoring when possible.

\paragraph{Ethics statement.}
This research was conducted in accordance with the NeurIPS Code of Ethics. No human subjects were involved. All models used are publicly available open-weight models. The total compute cost was under \$3 (OpenRouter API).

\paragraph{Conflict of interest.}
The author declares no competing interests.

% ── CONCLUSION ──
\section{Conclusion}

We present the first systematic investigation (to our knowledge) of whether scoring bias differs between base and instruct models. Our key finding---instruction tuning reduces format bias but may increase content bias---provides early evidence that scoring bias is modulated by training stage. The 22-model bias landscape (Table~\ref{tab:per_model}) offers the largest open comparison of scoring bias across models to date. We release all code and data for reproducibility and future extension.

All code, data, and analysis are publicly available at \url{https://github.com/ssamba1/Scoring-Bias-in-LLM-as-a-Judge}.

% ── Author Contributions ──
\section*{Author Contributions}
CRediT taxonomy: Conceptualization (lead), Methodology (lead), Software (lead), Validation (lead), Formal Analysis (lead), Investigation (lead), Data Curation (lead), Writing — Original Draft (lead), Writing — Review \& Editing (lead), Visualization (lead).

% ── PRE-REGISTRATION ──
\section*{Pre-registration}
This study was not pre-registered. Research questions (RQ1--RQ4) and hypotheses (H1--H4) were specified before data analysis. Post-hoc analyses are explicitly noted.

% ── ACKNOWLEDGMENTS ──
\section*{Acknowledgments}
We thank our research mentor for guidance and feedback. AI tools (large language models) were used to assist in code generation for data analysis and initial drafting of certain sections. All experimental design, data collection, analysis, interpretation, and editing were performed by the author.

% ── DATA AVAILABILITY ──
\section*{Data Availability}
All code and data are publicly available at \url{https://github.com/ssamba1/Scoring-Bias-in-LLM-as-a-Judge}. The experiment is fully reproducible at \$0 cost. A frozen snapshot of the code and data is archived at \href{https://doi.org/10.5281/zenodo.21361920}{10.5281/zenodo.21361920}.

\bibliographystyle{unsrt}
\bibliography{references}

\begin{appendix}
\section*{Supplementary Material}
\subsection*{A. Model Details}
\label{app:models}
Full model cards, including parameter counts, training methods, and licenses, are provided in \texttt{data/model\_cards/all\_models.md}.

\subsection*{B. Full Results}
\label{app:results}
Per-model bias scores, per-probe breakdowns, and item-level analysis are available at \url{https://github.com/ssamba1/Scoring-Bias-in-LLM-as-a-Judge}.

\subsection*{C. Formal Proofs}
\label{app:proofs}
The formal IIAR monotonicity theorem, differential direction corollary, and scaling law prediction are proved in \texttt{paper/supplementary\_formal.tex}.

\subsection*{D. Reproducibility Instructions}
\label{app:reproduce}
See \texttt{run\_all.sh} for the complete reproduction pipeline. Docker: \texttt{docker build -t scoring-bias . \&\& docker run scoring-bias}.
\end{appendix}

\end{document}
